The algebraic hierarchy in type theory using type classes

### 2010-06-22 jar.tex
Added formath
Add blurb about integers, rationals, and polynomials.

\subsection{Integers and rationals}

% The concrete implementation we use are just the Peano naturals \lstinline|nat|. Showing that \lstinline|nat| with its arithmetic operations and Leibniz equality form a \lstinline|SemiRing| is entirely routine, as is defining a map to an arbitrary other semiring and showing that this map is unique as a \lstinline|SemiRing_Morphism|, so that we get initiality.

@@

\paragraph{Acknowledgements.}
This research was stimulated by the new possibilities provided by the introduction of type classes
in Coq. In fact, this seems to be the first substantial development that tries to exploit all their
possibilities. As a consequence, we found many small bugs and unintended behavior in the type
class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
Georges Gonthier and Claudio Sacerdoti Coen have helped to
sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
%Jeremy, Thierry
\bibliographystyle{plain}
\bibliography{alg}
\end{document}

>>>>

\subsection{Integers, rationals, and polynomials}

The abstract interface for integers is completely analogous to the one for natural numbers:
\begin{lstlisting}
  Class Integers A {e plus mult zero one opp} `{IntegersToRing A} :=
    { integers_ring:> @Ring A e plus mult zero one opp
    ; integers_to_ring_mor:> Π `{Ring B}, Ring_Morphism (integers_to_ring A B)
    ; integers_initial:> Initial (ring.object A) }.
\end{lstlisting}

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and for which division of such integers is surjective:
\begin{lstlisting}
  Class Rationals A {e plus mult opp zero one mult_inv } `{Π x y: A, Decision (x = y)} {inj_inv}: Prop :=
    { rationals_field:> @Field A e plus mult opp zero one mult_inv
    ; rationals_frac: Surjective (λ p => integers_to_ring (Z nat) A (fst p) * / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinling|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinilne|Integers|, but doing so would provide no benefit.
  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
  % todo: say a few words about how we handle division by zero?

In our development we prove thas the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface. While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{something}. Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals. As stated earlier, we plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.

We also plan to do the exact same thing for polynomials, which can be abstractly specified as a free commutative algebra, and for which diverse implementations exist (coefficient lists, Bernstein polynomials, etc.).

@@

\paragraph{Acknowledgements.}
This research was stimulated by the new possibilities provided by the introduction of type classes
in Coq. In fact, this seems to be the first substantial development that tries to exploit all their
possibilities. As a consequence, we found many small bugs and unintended behavior in the type
class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
Georges Gonthier and Claudio Sacerdoti Coen have helped to
sharpen our understanding of the relation with Canonical Structures and with Unification Hints.

This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
%Jeremy, Thierry
\bibliographystyle{plain}
\bibliography{alg}
\end{document}
